%
% Introduction
%

\chapter{Introduction}
\label{c:Introduction}

\section{File Formats}

\subsection{User vs developer perspective}

File formats do often provide tough problems both for the
software engineers who write programs and for the people who are
using them.  Software engineers want formats that store data in
an efficient manner, and are easy to read and write. Users want
way to save their data in a convenient and fast manner, and don't
want to be bothered with the choice of a specific format.

\subsection{Converters}

The consequence is that almost every graphics or desktop publishing 
system has its own file format, optimized for the needs of that 
product. This means that direct data exchange between different 
products is impossible, since the file formats will be mutually 
exclusive. So, most programs contain lots of converters that 
transform data between different formats.

\subsection{The User's Perspective}

Having lots of converters this is inconvenient for the user. First, 
it means that $O(n^{2})$ converters are ideally needed to exchange data 
between n programs. However, it is unlikely that each program can 
read and write each other format.
If programs cannot exchange data since they don't understand each 
others format, but both understand a third program, there is a still 
a way to exchange data with that format. However, this is 
inconvenient because the user has to find out which programs are 
candidates for such an operation.
Furthermore, data may be lost in a format conversion. This may be 
because the other format is simply too complicated, or even secret 
information. There is obviously no way to avoid both issues, but they 
should not be as bothering as they are.

\subsection{The Developer's Perspective}

From an engineering point of view, it is conceivable that most
programs need their own format to represent their data in an
efficient way. However, it is not so obvious that the user needs
to be bothered with this.  First, consider converters. One easy
way to get rid of them is to provide one powerful format that has
a core part which is understood by all participating
applications, and can be extended to meet a particular format's
needs.  The RTF and SGML formats for marking text are such
approaches, The PICT graphics format is a successful example for
a graphics format.  Second, consider efficiency. It is often the
case that a more generalized format is less efficient in terms of
storage space or loading time. However, one can provide the
choice of saving data either in a native, efficient format, or in
a exchange format. Many desktop publishing programs use this
approach.

\section{Intermezzo: What is a graph ?}

\begin{definition}[Graph]
A graph is a tuple
\begin{eqnarray*}
  G & =         & (V,E), \mbox{\textnormal{where}} \\
  V &           & \mbox{\textnormal{is the set of \emph{nodes}, and}} \\
  E & \subseteq & V \times V \mbox{\textnormal{is the set of edges}}
\end{eqnarray*}
We also define a mapping \emph{label} which assigns information
to nodes, edges and labels:
\begin{displaymath}
  label: G \cup V \cup E \mapsto \Sigma^{*}
\end{displaymath}
\end{definition}

\begin{notes}
  \item In graph theory, $\Sigma$ is usually a fixed alphabet;
  GML needs a more general approach and allows to attach
  arbitrary attributes to graphs, nodes and edges.
\end{notes}


%
% A Simple GML Example
%

\section{A Simple GML Example}

\begin{example}%
{e:GML:Intro:circle3}%
{A simple graph in GML (circle of three nodes)}
\begin{alltt}
graph [                     \ttcomment{Defines a new graph}
    node [                  \ttcomment{Defines a new node}
        id 1                \ttcomment{This node has the id 1}
    ]
    node [                  \ttcomment{Defines a new node}
        id 2                \ttcomment{This node has the id 2}
    ]
    node [                  \ttcomment{Defines a new node}
        id 3                \ttcomment{This node has the id 3}
    ]
    edge [                  \ttcomment{Defines a new edge}
        source 1            \ttcomment{Source is the node with the id 1}
        target 2            \ttcomment{Target is the node with the id 2}
    ]
    edge [                  \ttcomment{Defines a new edge}
        source 2            \ttcomment{Source is the node with the id 2}
        target 3            \ttcomment{Target is the node with the id 3}
    ]
    edge [                  \ttcomment{Defines a new edge}
        source 3            \ttcomment{Source is the node with the id 3}
        target 1            \ttcomment{Target is the node with the id 1}
    ]
]
\end{alltt}
\end{example}

\begin{example}%
{e:GML:ComplexExample}%
{A complex GML example}
\begin{alltt}
# This file is in version 1 of GML
Version 1

graph [

  # \ttcomment{This graph has been created by the program "demo"}
  Vendor "demo"

  # directed \ttcomment{determines whether a graph is directed (1) or not (0).}
  # \ttcomment{In a directed graph, edges are have arrows that indicate the direction.}
  directed 1

  # \ttcomment{A label is a text attached to an object}
  label "The principles of space travel"

  node [  
    id 1
    label "Earth"
    graphics [
      x 0.1
      y 0.0
      w 0.1
      h 0.1
      image "earth.gif"
    ]
  ]

  node [
    id 2    
    label "Mars"
    graphics [
      x 0.9
      y 0.0
      w 0.055
      h 0.055
      image "Mars.gif"
    ]
  ]

  edge [
    source 1        
    target 2        
  ]
]
\end{alltt}
\end{example}

Example \ref{e:GML:Intro:circle3} shows a simple graph which
consists of three nodes.  The more complex example
\ref{e:GML:ComplexExample} demonstrates how to attach text to
graphs, nodes and edges, and how to specify coordinates and
images for nodes and edges.  These example illustrates the key
elements of GML:

\begin{itemize}

  \item A GML file is made up of pairs of a key and a value.
  Examples for keys are \texttt{graph}, \texttt{node} and
  \texttt{edge}.
  
  \item The key idea behind GML is that there are some
  standard keys like graph, node and edge, and anybody is free
  to add its keys to add specific information.

  \item Values can be integers, floating point numbers, strings
  and lists, where the latter must be enclosed in square
  brackets.
  
  \item The graph in Example \ref{e:GML:ComplexExample} did not
  specify how to place the labels.  They are arranged in a
  convenient manner in Figure 2, but an application might also
  ignore graph labels and print node labels over the images.
  There is no way to prevent a program from not drawing labels,
  but we will show in sections \ref{s:GML:NodeAttributes} and
  \ref{s:GML:EdgeAttributes} how the placement of labels is specified.

\end{itemize}



%
% Design Issues
%

\section{GML Design Issues}

\subsection{Syntax: Simple or Complex ?}

A complex syntax -- like a programming language -- gives the
designer the freedom to express facts in a efficient and
easy-to-read manner. The \texttt{dot} format is a good example
for this practice. However, the price to be paid is a less simple
implementation.  A simple syntax has the advantage that the
implementation is easier, but the format is less efficient in
terms of storage space and runtime. However, there is no reason
why the format should be less powerful; everything that can be
expressed with a complex format can be expressed in a simple
format.

\textbf{Answer: Simple} We chose a simple format over a complex
one. We loose some efficiency, but we gain a much easier
implementation and thus a wider distribution.  However,
simplicity is limited: since the format needs to be universal,
some details will be slightly complex. For example, strings are
always terminated by " characters. Therefore, we need a mechanism
to deal with a quote that appears inside a string. Other issues
are maximum line length and non ASCII characters, such as German
umlauts, \"{a}.


\subsection{Data types}

Which types of values do we need to represent? The answer is: all
data types present in programming languages. This includes
numbers (both integers and floating point), boolean values,
characters, strings and composite data types such as record,
array, set and list structures.


\subsection{Constraints}

There are several external constraints which have to be
considered:

\begin{description}
  
  \item[Maximum line length] Some systems cannot handle arbitrary
  long lines without problems.  Therefore, we need to restrict
  line lengths to a size that fits all systems.
  
  \item[Character Set] Internationalization is an important issue
  these days, and the ASCII characters are no longer sufficient
  for a real application.  Therefore, we will use the ISO 8859
  character set, which is a common way to code non ASCII
  character sets within ASCII.  ISO 8859 is also used in HTML, so
  we are in good company.
  
  \item[Range of values] The range of numbers is another
  sensitive point. We will assume 32 bit signed integers and
  double precision floating point values. These should be
  supported by all current systems.
  
  This rules out other data types like unsigned integers and 64
  bit integers, but those could still be stores as strings and
  converted afterwards.  We will not assume a maximum length for
  strings, as this would be a restrictions for applications that
  store long texts in strings.

\end{description}


%
% Notes on our Notation
%

\section{Notes on our Notation}

We use a Pascal like notation with some object oriented
extensions\footnote{Graphlet's implemention of GML is in C++, but
  we feel that Pascal is better for for aesthetic reasons, and
  more people are familiar with Pascal than with the fine prints
  of C++.}. A GML file is composed of key-value pairs, which we
call objects.  An object is a parametrized data type:

\begin{alltt}
type Object(\Param{Type}) = record
    key:   Key;
    value: \Param{Type};
end;
\end{alltt}

\noindent In the following, we will use the types \texttt{Object(Integer)},
\texttt{Object(Real)}, \texttt{Object(String)} and
\texttt{Object(List(Object))}.  We will also assume that the type
of an object is available at runtime, as in the following
example:

\begin{alltt}
case type(\Param{o}) of
    Object(Integer):      {\textnormal{\emph{Action for type}}} Integer;
    Object(Real):         {\textnormal{\emph{Action for type}}} Real;
    Object(String):       {\textnormal{\emph{Action for type}}} String;
    Object(List(Object)): {\textnormal{\emph{Action for type}}} List(Object);
end
\end{alltt}


%
% Paths
%

\subsection{Paths}

The data structure Object forms a tree, where elements of type
\texttt{Object(Integer)}, \texttt{Object(Real)} and
\texttt{Object(String)} are leaves, and elements of type
\texttt{Object(List(Object)))} are inner nodes.  We will
frequently use paths in this tree to describe the location of an
object.

\begin{definition}[Path]
  Let $k_{1},k_{2},\ldots,k_{n}, n \ge 1$ be keys. The path
  \[
  .k_{1}.k_{2}.\ldots.k_{n}
  \]
  denotes all sequences of objects $o_{1},o_{2},\ldots,o_{n}$ where
  \begin{eqnarray*}
    1 \le i \le n   & : & o_{i}.\mbox{key} = k_{i} \\
    1 \le i \le n-1 & : & \mbox{type}(o_{i}.\mbox{value}) =
    \mbox{Object(List(Object))} \\
    2 \le i \le n   & : & o_{i} \in o_{i-1}.\mbox{value}
  \end{eqnarray*}
\end{definition}

\noindent Examples of paths are \texttt{.id}, \texttt{.graph.label},
\texttt{.node.label} and \texttt{.node.graphics}. A path
describes a class of sequences of objects which can start
anywhere in the object tree. We can also define paths that start
at a specific object:

\begin{definition}[Path Starting at an Object]
  Let $k_{1}.k_{2}.\ldots.k_{n}, n \ge 1$ be a path and $o$ be an
  object. Then
  \[
  \mbox{\texttt{o}}.k_{1}.k_{2}.\ldots.k_{n}
  \]
  denotes the path starting at o, that is all sequences of
  objects $o_{1},o_{2},\ldots,o_{n}$ where
  \begin{eqnarray*}
    1 \le i \le n   & : & o_{i}.\mbox{key} = k_{i} \\
    1 \le i \le n-1 & : & \mbox{type}(o_{i}.\mbox{value}) = 
    \mbox{List(Object)} \\
    2 \le i \le n   & : & o_{i} \in o_{i-1}.\mbox{value} \\
    & & o_{1} \in o.\mbox{value}
  \end{eqnarray*}
\end{definition}

\noindent Finally, we omit the leading point on a path if it
starts at the root object:

\begin{definition}[Path Starting at the Root]
  Let $k_{1},k_{2},\ldots,k_{n}, n \ge 1$ be keys.
  \[
  k_{1}.k_{2}.\ldots.k_{n}
  \]
  denotes the path
  \[
  R.k_{1}.k_{2}.\ldots.k_{n}
  \]
  where $R$ is the root object of the tree.
\end{definition}


%
% Other Graph File Formats
%

\chapter{Other Graph File Formats}

\begin{note}
  This chapter is not complete yet.
\end{note}

\section{Simple adjacency lists}

Many systems use simple adjacency lists, perhaps enriched with labels 
or coordinates. Often, an adjacency list is terminated by the end of 
the line.
While this format type is convenient and easy to use in these 
systems, it has several disadvantages for our purpose. First, it is 
not expandible. Second, labels are usually restricted to one 
character or a single word. Further, the degree of a node is limited 
on systems which do not support arbitrary line lengths.

\section{GraphEd}

GraphEd had a format which is in spirit very simpliar to the one 
which is presented in this paper. However, its syntax is more complex 
than necessary in several aspects.

\begin{verbatim}
GRAPH "" =
1 {$ NS 32 32 $} ""
 2 ""
;
2 {$ NS 32 32 $} ""
 3 ""
;
3 {$ NS 32 32 $} ""
 1 ""
;
END
\end{verbatim}

This format contains several complications

\begin{itemize}

\item There are several ways to represent lists, e.g \verb|"[ Š]"|, 
\verb|"{$ Š $}"| and \texttt{GRAPH}\ldots\texttt{END}.

\item Adjacency lists start with a node and end with \texttt{";"}, 
whereas the list of nodes in a graph is terminated by \texttt{"END"}.

\item Some syntax elements are superficial, like the \texttt{"="} 
after the keyword \texttt{GRAPH}.

\end{itemize}

On the other hand, the format supported generic attributes (inside 
\verb|"{$ Š $}"|) which were very similar to the one we propose. The main 
difference was that GraphEd's attributes had a key and a list of 
values, where we will only have one value per key. Of course 
GraphEd's approach made it easier to write files, but the data 
structures behind that were more complex, since two list structures 
where needed. GraphEd needed a list of attributes and a list of 
values, whereas we need only a list of key-value pairs.
Another difference is that GraphEd put the graph structure and labels 
into a syntax outside the attributes, where we will combine them, 
once again to have only one data structure.

\section{TEI (SGML) Format}

The format used in TEI [reference] is actually a SGML DTD. It has the 
advantage that SGML provides a powerful standardized framework for it.
However, the SGML syntax is more complex than ours, so parsing is 
more difficult. Our goal is a format that can easily be converted 
into any other format, so a easy parser is essential. Also, we do not 
need a format with the power of SGML here.

\section{VRML Format}

VRML is a format who's syntax is quite similar to the one defined 
here. Basically, it uses a key-value structure. The main difference 
is that a key can have a list values.
The format is extensible.

\section{dot Format}
\TBD{}

\section{Tom Sawyer Software Format}

Tom Sawyer Software, Berkeley makes commercial graph layout and graph 
editor toolkits.  Their file format uses keys in a line started by 
\texttt{//} , followed by a list of values, each on its line:

\begin{verbatim}
// Graph Layout Toolkit
Hierarchical Layout
// minimumSlopePercent
20
// Nodes
2
42
\end{verbatim}

There is no hierarchical structure, although they can be modelled 
with dummy begin/end keys. The format is extensible; new elements can 
be added through a C or C++ interface.


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "GML"
%%% End: 
